A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2024-11-05 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: Intro and first section review

\section{Introduction}
Paper for physicists:
* argue that we need to develop mathematical structures from physical requirements
* create a theory of probability that works for classical, quantum mechanics and beyond
* inverted structure

@@

@@

Note that there may good conceptual grounds to require that a physical theory allows for statistical descriptions. If a physical theory is to be repeatedly testable experimentally, then we fail to see how it could be so without providing statistical tools. Every state preparation and measurement, in fact, is typically statistical in nature. Moreover, physical laws are statements of the type ``whenever we prepare this, we can measure that.'' A philosophical exploration of this area would be welcome.

We have identified three basic requirements on ensembles: first, they have to be identifiable experimentally; second, they need to allow mixtures; lastly, they need a well-defined entropy. Let us go through these items one by one.

\subsection{Experimental verifiability}
The first requirement is that ensemble must be connected to experimental verification. That is, we must have enough \textbf{verifiable statements} at our disposal to define ensembles and tell them apart. By verifiable statement we mean a statement for which an experimental test is available that terminates in finite time if and only if the statement is true. For example, the statement ``the mass of the photon is less than $10^{-13} \, eV$'' is verifiable, while ``the mass of the photon is exactly $0 \, eV$'' is not verifiable due to its infinite precision. Mathematically, this requirement forces an ensemble space to be a $T_0$ second countable topological space, where open sets represent verifiable statements and Borel sets represent testable statements without guarantee of termination.

The topological structure, then, represent the most foundational structure for a physical theory. It tells us why functions are ``well-behaved,'' which simply means they have to be topologically continuous in the ``natural topology'' induced by this requirement. It tells why the Banach-Tarski paradox does not apply in physics since non-Borel sets are physically ill-defined as they are not connected to experimental verification. Similarly, it tells us that sets with cardinality greater than that of the continuum are physically not relevant, as they cannot be given a $T_0$ second countable topology.

\subsection{Statistical mixtures}
Another basic requirement of an ensemble space is ability to prepare a \textbf{mixture} of two ensembles. If $\ens[a], \ens[b] \in \Ens$ are two ensembles, and $p \in [0,1]$ a weight, we will find the $\ens = p \ens[a] + (1-p) \ens[b] \in \Ens$ that describes a process that selects the first ensemble over the second $p$ percent of the times. Mathematically, the ensemble space is endowed with a \textbf{convex structure}.

Ultimately, the convex structure is responsible for all linear structures we have in physics. Both in classical and quantum mechanics, the ensemble space is not just a convex set, but a convex subset of a vector space. In those cases, in fact, the convex structure is, in a sense, invertible. If we fix $\ens$, $\ens[a]$ and $p$ then, if it exists, there is only one $\ens[b]$ such that $\ens = p \ens[a] + (1-p) \ens[b]$. The ensemble space is \textbf{complement}, meaning that there is only one complement to a component of an ensemble.

However, we found no physical justification for this requirement, and we suspect that it may need to be relaxed in future physical theories.

\subsection{Entropy}
The last requirement is that each ensemble must have a well defined entropy. That is, there is a scalar function $S : \Ens \to \mathbb{R}$ that satisfies the following
\begin{enumerate}
	\item strictly convex: $S(p \ens[a] + (1-p) \ens[b]) - (p S(\ens[a]) + (1-p) S(\ens[b]) ) \geq 0$
	\item upper bounded: $S(p \ens[a] + (1-p) \ens[b]) - (p S(\ens[a]) + (1-p) S(\ens[b]) ) \leq - p \log p - (1-p) \log(1-p)$
\end{enumerate}
The first tells us that, during mixing, the entropy cannot decrease, and it stays the same if and only we are mixing an ensemble with itself. The second tells us that the most the entropy can increase is given by the choice between the two ensembles.

>>>>

\section{Introduction}

Our current understanding of probability is given in terms of the measure theoretic formalization due to Kolmogorov, and its various philosophical interpretations. However, probability in quantum mechanics, as it is widely known, is not Kolmogorovian, in the sense that it provides results that are not reproducible in standard probability spaces. It follows that our mathematical axioms for probability and their interpretations are insufficient for physical theories.

An alternative measure theoretic strategy, that has been successful is quantum mechanics, is the use of quasi-probability distributions: measures that are still unitary over the whole space, but can have negative values in small regions. However, these haven't resulted in an alternative axiomatic approach with suitable interpretations for at least, in our mind, three reasons. First, it is not clear how negative probability should be interpreted. Second, it is not clear how to characterize the space of valid quasi-probability distributions (e.g. there is no simple rule to specify which $\rho(q,p)$ are valid Wigner functions). Third, there are multiple options for quasi-probability distributions (e.g. Wigner function, Husimi distribution, Sudarshan–Glauber distribution). This lack of foundational clarity presents an additional problem: how do we know whether future physical theories will not require yet another extension?

The goal of this paper is to show that, in the same way that classical probability can be axiomatized through standard measure theory, quantum probability, and in fact any probability used a physical theory, can be axiomatized through non-additive (i.e. fuzzy) measure theory. We will define a generalized concept of probability that is ultimately based on physically motivated axioms, giving the physicist a clear operational motivation to our definitions, the mathematicians a formally well-defined mathematical structure and the philosopher a unified framework in which probability can be interpreted in the context of physical theories. We are not going to be able to develop a full theory here, as there are still a number of conceptual and technical challenges to be solved that will require work in physics, mathematics and philosophy. However, we will show that there are enough results that point in a good direction, and therefore make it reasonable to invest time and effort to work on the open problems. For those interested, the current state of the research will always be available in an open document that will be routinely updated as the theory matures.

In a nutshell, we will start by arguing that physical theory must at least be able to talk about statistical ensembles, and that ensembles must allow the preparation of statistical mixture. Given a target ensemble $\ens$ and a set of ensemble $A$, we define the fraction capacity of $A$ for $\ens$ as the biggest fraction of $\ens$ that can be constructed from another set of ensembles $A$. That is, the highest $p$ such that we can express our ensemble as $\ens = p \ens[a] + (1-p) \ens[b]$, where $\ens[a]$ is a mixture of elements of $A$. For example, if $\ens$ is the uniform distributions for a roll of a six-faced die, and $A$ includes only two ensembles with $100\%$ probability for the outcomes $3$ and $4$ respectively, then the faction capacity of $A$ for $\ens$ will be $\frac{1}{3}$. Note that this coincides with the probability of the event ``$3$ or $4$''. The fraction capacity is a non-negative, unit bounded, sub-additive and continuous set function, which is therefore like a probability measure except for the additivity, which is recovered for classical mechanics and over quantum measurements.

\section{Quick review of probability theory}

@@

As mentioned in the introduction, our goal is to construct a probability theory that is guaranteed to work in any physical theory. Therefore our fundamental axioms will not be about probability, but about physical theories. This will help us place probability in a bigger context, and provide a minimal interpretation. As it is common in our larger project, we want axioms to describe necessary requirements of the scientific practice. That is, they will describe features of physical practice, not feature of the systems being described.

@@

At least two arguments can be made to show that any physical theory must provide a set of ensembles. First, note that physical laws are never about specific instances but rather regularities. They are statements of the type ``whenever we prepare this, we can measure that.'' This means that, ultimately, they are about ensembles. Second, if a physical theory is to be testable experimentally, and repeatedly so, it must be in terms of ensembles because this is what we prepare in practice and this is what, in the end, we characterized experimentally. It seems, then, that the approach can be given firm conceptual grounds, which we leave as open but solvable philosophical problem.

We have identified three basic requirements on ensembles: first, they have to be identifiable experimentally; second, they need to allow statistical mixtures; lastly, they need a well-defined entropy. Let us go through these items one by one. We leave the full details to the working document (TODO: decide how to call it).

\subsection{Experimental verifiability}
The first requirement is that ensemble must be connected to experimental verification. That is, we must have enough \textbf{verifiable statements} at our disposal to define ensembles and tell them apart. By verifiable statement we mean a statement for which an experimental test is available that terminates in finite time if and only if the statement is true. For example, the statement ``the mass of the photon is less than $10^{-13} \, eV$'' is verifiable, while ``the mass of the photon is exactly $0 \, eV$'' is not verifiable due to its infinite precision. Based on previous work, this mathematically requires an ensemble space to be a $T_0$ second countable topological space, where open sets represent verifiable statements and Borel sets represent statements associated with a test, but without guarantee of termination.

The topological structure, then, represent the most foundational structure for a physical theory. It tells us why functions are ``well-behaved,'' which simply means they have to be topologically continuous in the ``natural topology'' induced by this requirement. It tells why probability is assigned to Borel sets (i.e. all statements associated with tests) and why the Banach-Tarski paradox does not apply in physics (i.e. non-Borel sets are physically ill-defined as they are not connected to experimental verification). Similarly, it tells us that sets with cardinality greater than that of the continuum are physically not relevant, as they cannot be given a $T_0$ second countable topology.

\subsection{Statistical mixtures}
Another basic requirement is the ability to prepare a \textbf{mixture} of two ensembles. If $\ens[a], \ens[b] \in \Ens$ are two ensembles, and $p \in [0,1]$ a weight, then $\ens = p \ens[a] + (1-p) \ens[b] \in \Ens$ that describes a process that selects the first ensemble over the second $p$ percent of the times. Mathematically, the ensemble space is endowed with a \textbf{convex structure}.

Ultimately, the convex structure is responsible for all linear structures we have in physics. Both in classical and quantum mechanics, the ensemble space is not just a convex set, but a convex subset of a vector space. In those cases, in fact, the convex structure is, in a sense, invertible. If we fix $\ens$, $\ens[a]$ and $p$ then, if it exists, there is only one $\ens[b]$ such that $\ens = p \ens[a] + (1-p) \ens[b]$. The ensemble space is \textbf{complemented}, meaning that there is only one complement to a component of an ensemble.

This perspective allows to understand what negative probability is. Suppose we have a real vector space of dimensions $n$, then $n$ linearly independent vectors are enough to express every point as a linear combination. If we add another point, we can now express every point as an affine combination, that is a linear combination where the coefficients sum to one, though some can still be negative. Since the space of mixed space in quantum mechanics is a real vector space, we can choose a linearly dependent set of states space span the whole space, and every mixed state can be expressed as an affine combination, as a pseudo-probability.

Pseudo-probability distributions, then, require the additional vector space structure. However, we found no physical justification to require the additional structure, and we suspect that it may need to be relaxed in future physical theories.  Therefore it is important that our fundamental definitions require only the convex structure, and not the vector space structure.

\subsection{Entropy}
The last requirement is that each ensemble must have a well defined entropy. That is, there is a scalar function $S : \Ens \to \mathbb{R}$ that satisfies the following
\begin{enumerate}
	\item strictly convex: $S(p \ens[a] + (1-p) \ens[b]) - (p S(\ens[a]) + (1-p) S(\ens[b]) ) \geq 0$
	\item upper bounded: $S(p \ens[a] + (1-p) \ens[b]) - (p S(\ens[a]) + (1-p) S(\ens[b]) ) \leq - p \log p - (1-p) \log(1-p)$
\end{enumerate}
The first tells us that, during mixing, the entropy cannot decrease, and it stays the same if and only we are mixing an ensemble with itself. The second tells us that the most the entropy can increase is given by the choice between the two ensembles.\footnote{The use of the Shannon entropy for the upper bound can likely be recovered.}
